---
title: "This is the Vignette for Finalfunctions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{This is the Vignette for Finalfunctions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<font size="6">Final Project: A R Package For Electrophysiology Data Analysis</font>
<br/>
<font size="6">Purpose of the Package</font>  
<font size="3">This package is specifically designed to analyze electrophysiology data from an experiment in my lab. Basically, the experiment is about investigating how chronic alcohol drinking alters the animals' brain circuits (Motor Cortex --> Dorsolateral striatum). This circuit is known to play a role in habit formation and involved in addiction. We need to compare the cells' voltage threshold and current threshold for example, from alcohol drinking animals with water control animals. The lower the threshold, the easier to fire an action potential, the more excitatory connections between the circuits.   
Without making this package, there are a lot of works need to be done to analyze the data. First, manually exclude the cells that are not stable based on several criteria. Second, use excel to reorganize the sheets and calculate mean, STD, n and some other basic properties for each variable. Third, import all the data to Prism, reorganize them, perform statistical analysis, make graphs, and make them pretty. We need to do all the above steps over and over again every time we get new data, which is a lot of repetitive work. In this case, this package is designed to make things a lot easier. It contains several functions that are specifically designed to do all of the above steps required. This can quickly do data analysis and graph plotting, which would save a lot of time. The general processes include:   
1. Filter out unstable/invalid cells based on the value of their y frac, z plane, and resistance differences.  
2. Plot boxplot/violin plot for voltage threshold and current threshold data.  
3. Perform t-test analysis on voltage threshold and current threshold (saline group vs. alcohol group).     
Overall, this project will highly improve the efficiency for data analysis on this experiment. The following will be the explanation of each function and some examples on how to use them. </font>  


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(FinalProjectYin)
library(dplyr)
library(knitr)
library(kableExtra)
library(flextable)
library(ggplot2)
library(ggpubr)
```

<font size="5">Read the dataset for example demonstration.</font>    
<font size="3">I used one of the electrophysiology datasets from my lab. The file has information including cell ID, their treatment groups (either alcohol or water), y frac value, z plane value, start and end resistance value, percent resistance difference, voltage threshold (mV), and current threshold (pA).</font>    
```{r}
ephys <- read.csv("Ephys_Yin.csv", header = TRUE)
```


<font size="5">Function 1: ephysfilter</font>         
<font size="3">This function helps with the first step of data analysis, which is filtering data. There are some criteria regarding the cell quality. The has to be:        
1. The y frac value should fall between 0.4-0.6.      
2. The z plane value should fall between 35-75.      
3. Resistance difference should be less than 20%, which is less than 0.2 from the table.    
The following is an example: By simply putting the original file name into this  \func{ephysfilter()}, we can easily filter out those unstable cells that do not fit those criteria and put the filtered data into a new file called ephys1.
</font>
```{r}
head(ephys1 <- ephysfilter(ephys))
```

<font size="5">Function 2: Thresholdplot</font>        
<font size="3">The second step is to plot a graph to visualize the data. I decided to use violin graph with a box plot in it because violin plot can not only show a marker for the median of the data, a box indicating the interquartile range, but also provide additional information about the probability density of the data at different values.Though we need to install the ggplot2 package in order to use this function, it still saves a lot of time by not having to type in all the labeling, captions, and requirements.    
The following is an example: But puting the dataname, treatment group, dependent variable name, and the y-axis title into the \func{Thresholdplot()}, it will generate a violin plot: The a-axis is the treatment group (alcohol and water). The y-axis is the dependent variable, which is voltage threshold (mV) in this example. The title above is called "Voltage Threshold (mV)".</font>   
```{r}
Thresholdplot(ephys1, "Treatment", "Voltage.threshold.mV", "Voltage Threshold (mV)")
```

<font size="5">Function 3: Create a Statistical Table</font>    
<font size="3">After visualizing the data, we need to calculate the statistics to see if the two groups are actually significantly different from each other. This \function{StatsTable} helps with creating a statistical table which shows the mean and standard deviation of the two treatment groups. A couple of steps were included in this function:    
1. Create a data frame with only one treatment group.    
2. Select only the cell ID and the dependent variable columns in that data frame.    
3. Calculate the mean and standard deviation for the selected dependent variable.    
4. Repeat the above steps for the other treatment group.    
5. Combine those two data frame into one.    
This \function{StatsTable} helps save a lot of steps and time to get to this statistical table. The following are two examples of how to use this function. By putting the data name, two treatment groups' name, and the name of the dependent variable, we can easily get the mean and STD for alcohol and water groups' voltage threshold and current threshold.</font>
```{r}
StatsTable(ephys1, "alcohol", "water", "Cell.ID", "Voltage.threshold.mV")
StatsTable(ephys1, "alcohol", "water", "Cell.ID", "Current.Threshold..pA.")
```

<font size="5">Function 4: Visualizing Data Normality</font>    
<font size="3">Before performing the t-test for the above two treatment groups, we need to determine if the data of each group is normally distributed. To visualize the data normality, we can use Quantile-quantile plots (Q-Q plots), which draws the correlation between a given sample and the normal distribution.If most of the data points are fitted on the line, that means the data set is normally distributed.        
The following is an example of alcohol treated group's voltage threshold data distribution. By inputting the file name, treatment group name, the dependent variable name, and the title you wanted, a Q-Q plot will be generated with a linear regression line. We can see that all the data points are close to the line, indicating that all data can be explained by this model. The data set is likely normally distributed based on the graph.</font>
```{r}
normaldisplot(ephys1, "alcohol", "Voltage.threshold.mV", "Voltage Threshold (mV)")
```


<font size="5">Function 5: Check Data Normalization</font>    
<font size="3">Besides visualizing the distribution of the data, we have to calculate the statistical results to prove if the data set is actually normally distributed. This \function{normalization()} helps with two types of preliminary tests to check unpaired two-sample t-test assumptions.    
The first one is to use Shapiro-Wilk normality test to look at the normality. If the p-value is larger than 0.05, it means the distribution of the data are not significantly different from normal distribution. So we can assume the normality.    
The second one is to use F-test to test for homogeneity in variance by using var.test(). If the p-value is greater than 0.05, it means there's no significant difference between the variance of the two sets of data. So we can use the classic t-test for the analysis.    
Below is an example of the preliminary tests of the voltage threshold data in alcohol treated group. After providing the data name, dependent variable, and the treatment group in the function, we can quickly get the results of the two tests: (1) Shapiro-Wilk normality test p-value=0.1962, which is larger than 0.05. This means the alcohol group data has the normality. (2) F-test p-value=0.6951, which is also larger than 0.05. It indicates that alcohol group data has homogeneity in variance. We can move on to use the classic t-test for analysis.</font>  
```{r}
normalization(ephys1, "Voltage.threshold.mV", "alcohol")
```

<font size="5">Function 6: T-Test</font>    
<font size="3">The final step is to perform a t-test to study if there is a significant difference between the two treatment groups. By simply typing in the dependent variable of interest and the file name, we can get the p value of the t-test and know the relationship between these two groups.    
The following are two examples of using this function. The results showed that the p-value for voltage threshold is 0.01275359, which is smaller than 0.05. This indicates that there is a significant difference of voltage threshold between alcohol and water groups. However, the p-value for current threshold is 1, which is larger than 0.05. It means there's no difference in current threshold between these two groups.</font>
```{r}
threshttest("Voltage.threshold.mV", ephys1)
threshttest("Current.Threshold..pA.", ephys1)
```

<font size="5">Conclusion</font>    
<font size="3">Overall, this package helps make the data analysis of electrophysiology data in our lab a lot easier and faster. As long as we keep organizing our raw data in a similar pattern, by using the above six functions, we can quickly filter the data, draw beautiful graphs, and get statistical results.</font>
